### 2022-10-22 Papers/ICCC-2023/sections/abstract.tex
corrections

As static metaprogramming is becoming more relevant, compilers must adapt to acomodate them.
This requires exposing more information about the code, from the compiler, to the programmer as well as more powerfull compile-time function execution capabilities.
The interpreter component of a compiler therefore becomes more and more prominent.

In this paper a new approach to compiler architecture that places the interpreter as the central component of the compiler.
Translation of user code into the executable form is done using an interpreted module, written in the target language, and uses the same data structures accessible to the programmer.
This boosts the flexibility and extensibility of the compiler and ensures the completness of the available metadata and the ability to execute any code at compile-time.

>>>>

As static metaprogramming is becoming more relevant, compilers must adapt to acomodate them.
This requires exposing more information about the code, from the compiler, to the programmer as well as more powerfull compile-time function execution capabilities.
The interpreter component of a compiler therefore becomes more and more important.

In this paper a new approach to compiler architecture that places the interpreter as the central component of the compiler.
Translation of user code into the executable form is done using an interpreted module, written in the target language, and uses the same data structures accessible to the programmer.
This improves the flexibility and extensibility of the compiler and ensures the completness of the available metadata and the ability to execute any code at compile-time.

### 2022-10-22 Papers/ICCC-2023/sections/conclusions.tex
corrections

CTFE First approach to compiler construction offers high degree of compiler extensibility, at the cost of development time and performance.
Since user code has access to the same information as the compiler at compile time, it may perform more advanced functions.
The thesis that introduced this approach \cite{grabski2020} demonstrated numerous practical application: generating bindings for other languages, static analysis and extending semantics of the language.
These these goals are significantly easier to accomplish, thanks to the access to semantic model of the program, constructed by the compiler.
Authors of language tools do not need to analyze the user program.

>>>>

CTFE First approach to compiler construction offers high degree of compiler extensibility, at the cost of development time and performance, compared with conventional compilers.
Since user code has access to the same information as the compiler at compile time, it may perform more advanced functions.
The thesis that introduced this approach \cite{grabski2020} demonstrated numerous practical application: generating bindings for other languages, static analysis and extending semantics of the language.
These goals are significantly easier to accomplish, thanks to the access to semantic model of the program, constructed by the compiler.
Authors of language tools do not need to analyze the user program.

### 2022-10-22 Papers/ICCC-2023/sections/design.tex
corrections

C-=-1 was designed as a compiled, low-level, non-garbage-collected programming language, similar to C, C++ or Rust.
What diffirentiates C-=-1 are its two founding principles: all code is executable at compile-time and support rich metaprogramming.
The primary purpose of the language was to research how these ideas influence software written in it \cite{grabski2022compilation}.

C-=-1 is a simple language, built with the minimum set of features needed to demonstrate the usefulness of the proposed metaprogramming features.
They are discussed further in chapter \ref{design:attributes_and_metaprogramming}.
The primary motivation for those mechanisms, is providing the programmer with the ability to create domain specific static analysis and code generation tools, without creating a seperate program.

@@

C-=-1 type system heavily borrows from C++.
Program may contain user defined classes, with members which may have limited accessibility.
Generic programming is achieved by the use of templates, although they are much more limited than the ones present in C++.
The user may also use pointers to objects, with arbitrary indirection (for example pointer to pointer to object).
Additionally, the language contains the concept of an \lstinline{interface}, similar to the one found in C\#.

@@

Metaprogramming in C-=-1 is based around attributes.
Attributes work in a manner similar to the ones found in C\#.
They are types, which may contain fields and methods.
They can also, be used to annotate other elements of the program, such as types, functions or variables.

In C-=-1 these attributes may implement special member functions that react to uses of the annotated program element.
For example, an attribute that can be attached to a function, may implement \lstinline{onCall} special member function.
It will be called, at compile time, for each invocation of the annotated procedure.
Within the special method, the attribute will have access to the semantic model of the call site.
It may then modify the semantic model or report warnings or errors.

@@

To declare an attribute in C-=-1, the \lstinline{att} keyword is used.
After that, attribute targets should be listed in angled brackets, as on line 0 of Listing \ref{lst:noDiscardCm1}.
Attributes may target any number of language elements, including types, functions, variables and fields.
Attribute from Listing \ref{lst:noDiscardCm1} declares two member functions: \lstinline{attach} on line 4 and \lstinline{onCall} on line 6.

The \lstinline{attach} method is called after names of all of the program elements have been gathered, but before the compiler started to analyze function bodies.
It is common accross all attribute targets and accepts the descriptor of the attached program element (function, field, type, etc.).
Only during call to \lstinline{attach} can the attribute change aspects of the program that affect function overload resolution, for example
whether a function is invokable at run or compile time.

@@

CTFE First apporach was created during implementation of the first compiler for the C-=-1 language\cite{grabski2022compilation}.
CTFE First compiler has four major components: Frontend, Interpreter, Compiler Interface and Backend.
Figure \ref{CTFE-first-compiler-structure} contains a diagram with an overview of how these parts interact with eachother, during the compilaton process.
Frontend, described in chapter \ref{frontend}, parses the code in the compiled language and constructs its intermidaite representation, using interpreter's data structures.
It is used to analyze both user code and the Compiler Interface.
After the intermidiate representation is constructed, it is passed on to the Interpreter, which was described in chapter \ref{interpreter}.
Compiler Interface intermidiate representation is then executed, using the user program as data.
This step converts the semantic model of the program into the backends intermidiate language.
This process is further explained in chapter \ref{compiler-interface}.
Finaly the Backend generates the executable file.

@@

In the CTFE First approach, frontend serves the same role of constructing the programs intermidiate representation, as in conventional compilers \cite{puntambekar:compiler_design}.
The major difference lays in the data structures used to represent the program.

@@

In CTFE First, interpreter is the heart of the compiler.
It executes the Compiler-Interface which translates the intermidiate representation into the backend's assembly and serves as what is sometimes called the 'middle-end' of the compiler\cite{hsu2021llvm}.
To do it, it must be able to treat the program's intermidiate representation both as code and data.

@@

Compiler Interface translates the program's intermidaite representation into the backend's assembly language.
This component is interpreted during compilation.
What is unique about CTFE First is that this part of the compiler can thus be written in the target language, during Stage 0 of the compiler bootstrapping process, as was the case for the C-=-1 compiler\cite{puntambekar:compiler_design, novillo2007gcc, grabski2022compilation}.

@@

The backend code must be invokable from within the interpreted program in the target language.
Depending on how the interpreter was designed, this may require significant effort. %todo: maybe other word
Compiler backends are large and for the Compiler Interface to take advantage of them, they must be fully available.
This means exposing each function and type within the library, to the interpreted code.
These bindings could feasibly be generated automatically \cite{marshalling_auto_generation}, but this tehnique was not used when implementing C-=-1 compiler.

>>>>

C-=-1 was designed as a compiled, low-level, non-garbage-collected programming language, similar to C, C++ or Rust.
What diffirentiates C-=-1 are its two founding principles: all code is executable at compile-time and support rich metaprogramming.
The primary purpose of the language was to investigate how these ideas influence software written in it \cite{grabski2022compilation}.

C-=-1 is a simple language, built with the minimum set of features needed to demonstrate the usefulness of the proposed metaprogramming features.
They are discussed further in chapter \ref{design:attributes_and_metaprogramming}.
The primary motivation for those mechanisms, is providing the programmer the ability to create domain specific static analysis and code generation tools, without creating a separate program.

@@

C-=-1 type system borrows from C++.
Program may contain user defined classes, with members which may have limited accessibility.
Generic programming is achieved by the use of templates, although they are much more limited than the ones present in C++.
The user may also use pointers to objects, with arbitrary indirection (for example pointer to pointer to object).
Additionally, the language contains the concept of an \lstinline{interface}, similar to the one found in C\# \cite{hejlsberg2003c}.

@@

Metaprogramming in C-=-1 is based on attributes.
Attributes work in a manner similar to the ones found in C\#.
They are types, which may contain fields and methods.
They can also, be used to annotate other elements of the program, such as types, functions or variables.

In C-=-1 these attributes may implement special member functions that react to use of the annotated program element.
For example, an attribute that can be attached to a function, may implement \lstinline{onCall} special member function.
It will be called, at compile time, for each invocation of the annotated procedure.
Within the special method, the attribute will have access to the semantic model of the call site.
It may then modify the semantic model or report warnings or errors.

@@

To declare an attribute in C-=-1, the \lstinline{att} keyword is used.
After that, attribute targets should be listed in angled brackets, as in line 0 of Listing \ref{lst:noDiscardCm1}.
Attributes may target any number of language elements, including types, functions, variables and fields.
Attribute from Listing \ref{lst:noDiscardCm1} declares two member functions: \lstinline{attach} in line 4 and \lstinline{onCall} in line 6.

The \lstinline{attach} method is called after names of all of the program elements have been gathered, but before the compiler starts to analyze function bodies.
It is common accross all attribute targets and accepts the descriptor of the attached program element (function, field, type, etc.).
Only during call to \lstinline{attach} can the attribute change aspects of the program that affect function overload resolution, for example
whether a function is invokable at run or compile time.

@@

CTFE First apporach was created during implementation of the first compiler for the C-=-1 language\cite{grabski2022compilation}.
CTFE First compiler has four major components: Frontend, Interpreter, Compiler Interface and Backend.
Figure \ref{CTFE-first-compiler-structure} contains a diagram with an overview of how these parts interact with eachother, during the compilaton process.
Frontend, described in chapter \ref{frontend}, parses the code in the compiled language and constructs its intermediate representation, using interpreter's data structures.
It is used to analyze both user code and the Compiler Interface.
After the intermediate representation is constructed, it is passed to the Interpreter, which was described in chapter \ref{interpreter}.
Compiler Interface intermediate representation is then executed, using the user program as data.
This step converts the semantic model of the program into the backends intermediate language.
This process is further explained in chapter \ref{compiler-interface}.
Finaly the Backend generates the executable file.

@@

In the CTFE First approach, frontend serves the same role of constructing the programs intermediate representation, as in conventional compilers \cite{puntambekar:compiler_design}.
The major difference lays in the data structures used to describe the program.

@@

In CTFE First, interpreter is main component of the compiler.
It executes the Compiler-Interface which translates the intermediate representation into the backend's assembly and serves as what is sometimes called the 'middle-end' of the compiler\cite{hsu2021llvm}.
To do it, it must be able to treat the program's intermediate representation both as code and data.

@@

Compiler Interface translates the program's intermediate representation into the backend's assembly language.
This component is interpreted during compilation.
What is unique about CTFE First is that this part of the compiler can thus be written in the target language, during Stage 0 of the compiler bootstrapping process, as in the case of the C-=-1 compiler\cite{puntambekar:compiler_design, novillo2007gcc, grabski2022compilation}.

@@

The backend code must be invokable from within the interpreted program in the target language.
Depending on how the interpreter was designed, this may require significant effort. %todo: maybe other word
Compiler backends are large and for the Compiler Interface to take advantage of them, they must be fully available.
This means exposing each function and type within the library, to the interpreted code.
These bindings could feasibly be generated automatically \cite{marshalling_auto_generation}, but this technique was not used when implementing C-=-1 compiler.

### 2022-10-22 Papers/ICCC-2023/sections/implementation.tex
corrections

\subsection{Interpreter data structures}
\label{data_structures}
Data structures of the C-=-1 Interpreter have been designed ease of development and debugging in mind.
They are thus not particularly efficent.

@@

There are multiple types of reference within the C-=-1 interpreter.
The most basic pointer type is a reference to C-=-1 value.
It was realised as a pointer to the owning pointer of the value.
The reasoning behind this decision was to allow assigning to the referenced value.

@@

A major motivation in creating CTFE First was the ability to support languages with compile-time metaprogramming.
This includes reflection and modification of the code being compiled.
User program has to be represented as a complete and modifiable object, using the interpreters data structures.

C-=-1 language model divides the user program into assemblies.
They represent an individual program package: a library or an executable file.
The compiler is invoked to compile an assembly, together with its dependencies.
Assembly is the root object of C-=-1 program model, it stores a list of assemblies it depends on and the root namespace of the package it represents.
The remainder of user code is organised around namespaces, types, functions and fields.
These parts of the model are repesented by native classes of the host language and form the basis for the rest of the model.

@@

The C-=-1 backend interface utylises LLVM \cite{llvmir} code generation API that has been exposed by the compiler.
The functionality which has been made available to C-=-1 represents a minimal subset of LLVM, that is sufficent to implementing a basic compiler.

@@

Listing \ref{impl:theoretical_llvmir} contains a simple function written in C-=-1 (line 2), its ideal LLVMIR (line 5) and LLVMIR generated by C-=-1 compiler (line 21).
The current implementation generates a function for each operator, regardless of whether it was defined by the programmer or is a prymitive operation.
They are merely a wrapper around the actual LLVM intrinsic, meant to simplify implementation of the Backend Interface.
Future versions, with additional effort, may generate the ideal LLVMIR from line 21 of Listing \ref{impl:theoretical_llvmir}.

@@

Certain other intrinsic operations were defined using external dependencies.
C-=-1 memory managment library, in the runtime context, uses a simple interface capable of allocating and deleting a continous buffer.
It is comprised of two functions: \lstinline{unsafe_new} and \lstinline{delete}.
In the standard library, they are explicitly mapped to \lstinline{malloc} and \lstinline{free} functions from the C runtime.

>>>>

\subsection{Interpreter data structures}
\label{data_structures}
Data structures of the C-=-1 Interpreter have been designed having the ease of development and debugging in mind.
They are thus not particularly efficent.

@@

There are several types of reference within the C-=-1 interpreter.
The most basic pointer type is a reference to C-=-1 value.
It was realised as a pointer to the owning pointer of the value.
The reasoning behind this decision was to allow assigning to the referenced value.

@@

A major motivation for creating CTFE First was the ability to support languages with compile-time metaprogramming.
This includes reflection and modification of the code being compiled.
User program has to be represented as a complete and modifiable object, using the interpreters data structures.

C-=-1 language model divides the user program into assemblies.
They represent an individual program package: a library or an executable file.
The compiler is invoked to compile an assembly, together with its dependencies.
Assembly is the root object of C-=-1 program model, it stores a list of assemblies it depends on and the root namespace of the package it represents.
The remainder of user code is organised into namespaces, types, functions and fields.
These parts of the model are repesented by native classes of the host language and form the basis for the rest of the model.

@@

The C-=-1 backend interface uses LLVM \cite{llvmir} code generation API that has been exposed by the compiler.
The functionality which has been made available to C-=-1 represents a minimal subset of LLVM, that is sufficent to implement a basic compiler.

@@

Listing \ref{impl:theoretical_llvmir} contains a simple function written in C-=-1 (line 2), its ideal LLVMIR (line 5) and LLVMIR generated by C-=-1 compiler (line 21).
The current implementation generates a function for each operator, regardless of whether it was defined by the programmer or is a primitive operation.
They are merely a wrapper around the actual LLVM intrinsic, meant to simplify implementation of the Backend Interface.
Future versions, with additional effort, may generate the ideal LLVMIR from line 21 of Listing \ref{impl:theoretical_llvmir}.

@@

Certain other intrinsic operations are defined using external dependencies.
C-=-1 memory managment library, in the runtime context, uses a simple interface capable of allocating and deleting a continous buffer.
It is comprised of two functions: \lstinline{unsafe_new} and \lstinline{delete}.
In the standard library, they are explicitly mapped to \lstinline{malloc} and \lstinline{free} functions from the C runtime.

### 2022-10-22 Papers/ICCC-2023/sections/introduction.tex
corrections

Compile-time function execution (CTFE) is an aspect of a programming language, that allows the programmer to execute code at compile-time.
It has been gaining popularity with low level programming langauges such as Rust or C++.
The goals and capabilities of CTFE depend on the language and are discussed in chapter \ref{related-work}, however it plays a secondary role in the compilation process.

In this paper a new approach to compiler construction that places Compile Time Function Execution capabilities as the first priority is proposed.
This architecture is called CTFE First.
CTFE First builds the compiler around the interpreter component.
The role of the compiler is to construct the semantic model of the program being compiled and pass it as data to the interpreted Compiler-Intreface module.
Compiler-Interface then tranforms the model into the backend's assembly language to be compiled to executable.
The design of the language that motivated this approach was described in chapter \ref{language-design}.
Chapter \ref{compiler-design} deals with structure of a CTFE First compiler and how its components interact.

>>>>

Compile-time function execution (CTFE) is an aspect of a programming language, that allows the programmer to execute code at compile-time.
It has been gaining popularity with low level programming langauges such as Rust \cite{rust} or C++ \cite{ISO:cpp98}.
The goals and capabilities of CTFE depend on the language and are discussed in chapter \ref{related-work}.

In this paper a new approach to compiler construction that places Compile Time Function Execution capabilities as the first priority is proposed.
This architecture is called CTFE First.
CTFE First builds the compiler around the interpreter component.
The role of the compiler is to construct the semantic model of the program being compiled and pass it as data to the interpreted Compiler-Intreface module.
Compiler-Interface then tranforms the model into the backend's assembly language to be compiled to executable.
The design of the language that motivated this approach is described in chapter \ref{language-design}.
Chapter \ref{compiler-design} describes the structure of a CTFE First compiler and how its components interact.

### 2022-10-22 Papers/ICCC-2023/sections/realted-work.tex
corrections

Rust language compiler is the most similar in capability, to what was demonstrated with C-=-1, as a feature of CTFE First.
It allows the user to write code that performs some transformations of the program, and interact with the evnoirment during the build process.
The two relevant features are build scripts and macros.

@@

Build scripts, on the other hand, are programs that execute prior to compilation of the main package.
They preparing the envoirment for building the program by, for example, compiling its external dependencies.
They do not have any information about the structure of the program being compiled.
If a task, done by a build script, were to require such data, it would have to analyze the code by itself, without any help from the compiler.
This makes application such as automatically genertaing bindings for other programming languages significantly more difficult.

>>>>

Rust language compiler is the most similar in capability, to what was demonstrated with C-=-1, as a feature of CTFE First.
It allows the user to write code that performs some transformations of the program, and interact with the envoirment during the build process.
The two relevant features are build scripts and macros.

@@

Build scripts, on the other hand, are programs that execute prior to compilation of the main package.
They prepare the envoirment for building the program by, for example, compiling its external dependencies.
They do not have any information about the structure of the program being compiled.
Build scripts that require such data, would have to analyze the code by itself, without any help from the compiler.
This makes application such as automatically genertaing bindings for other programming languages significantly more difficult.
